---
title: "Chapter 3"
author: "Dimitrios Tsallos"
date: "11/21/2017"
output: html_document
---

# 3^nd^ Week

## Logistic Regression

```{r, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
# set working directory
setwd( "/Users/dtsallos/Desktop/OneDrive - University of Helsinki/GitHub/IODS-project/")
# read the alc data into memory
alc <- read.csv("data/student-combined.csv", sep = ",",header=TRUE)
alc <- alc[,-1] #remove column X with row numbers
```

### Student performance dataset

This week, we are working with a new dataset. This data approach student achievement in secondary education of two Portuguese schools. The data attributes include student grades, demographic, social and school related features) and it was collected by using school reports and questionnaires. Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). In [Cortez and Silva, 2008], the two datasets were modeled under binary/five-level classification and regression tasks. 

high and low alcohol consumption should be mentiond...

*You can find out more about this dataset and the related study [here](https://archive.ics.uci.edu/ml/datasets/Student+Performance).* 

But here is the variable names included in the following analysis: 
  
  
```{r, echo=FALSE}
colnames(alc)
```

And here is how the data has been structured:

```{r, echo=FALSE}
str(alc)

```


## Hypothesis

Using the given data, we are going to study student's behavior with regards to alcohol consumption. According to stereotypes, students who are being raised by one of the two parents alone (Pstatus = A, "apart") are more likely to seek alcohol to soothen emotional distress. When those students are members of big families (famsize = GT3, "greater than 3") the situation could lead to higher alcohol consumption. These students are also more likely to miss class (absences are numeric from 0 to 93). While those students who are busy with extracurricular activities (activities = yes) could be distracted, get more supervision and support which leads to less alcohol consumption and less abcences recorded from class. Let's study now if that is true based on the given data sample.  

famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3) 
Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart) 
activities - extra-curricular activities (binary: yes or no) 
absences - number of school absences (numeric: from 0 to 93) 


## Visualise data trends 

Numerically and graphically explore the distributions of your chosen variables and their relationships with alcohol consumption (use for example cross-tabulations, bar plots and box plots). Comment on your findings and compare the results of your exploration to your previously stated hypotheses. (0-5 points)


Comments missing...

```{r, echo=FALSE}
library(dplyr)
library(gmodels)
CrossTable(alc$high_use, alc$famsize, dnn = c("High alcohol use", "Family size"))
CrossTable(alc$high_use, alc$Pstatus, dnn = c("High alcohol use", "Parental status"))
CrossTable(alc$high_use, alc$activities, dnn = c("High alcohol use", "Extra-curricular activities"))

```




```{r, echo=FALSE}
library(ggplot2)
p1 <- ggplot(alc, aes(x= absences, fill=high_use)) + geom_density(alpha=.3)

p1 + ggtitle("Absence distribution") + 
    theme_bw() + 
    annotate("text", x = 16, y= 0.14,
                 label= paste( "High alchol Average =", round( mean( alc[ which(alc$high_use == TRUE), 30  ] ), digits = 2 ) ),
             vjust=1.5, size=6) + 
    annotate("text", x = 15, y= 0.125,
                 label= paste( "High alcohol Median =", median( alc[ which(alc$high_use == TRUE), 30  ] )  ), 
             vjust=1.5, size=6)

```


### Logistic Regretion


Use logistic regression to statistically explore the relationship between your chosen variables and the binary high/low alcohol consumption variable as the target variable. Present and interpret a summary of the fitted model. Present and interpret the coefficients of the model as odds ratios and provide confidence intervals for them. Interpret the results and compare them to your previously stated hypothesis. Hint: If your model includes factor variables see for example the first answer of this stackexchange thread on how R treats and how you should interpret these variables in the model output (or use some other resource to study this). (0-5 points)


```{r, echo=FALSE, message=FALSE}
# find the model with glm()
m <- glm(high_use ~ famsize + activities + Pstatus + absences, data = alc, family = "binomial")

# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI <- confint(m) %>% exp

# print out a summary of the model
summary(m)

# print out the coefficients of the model
coef(m)


# print out the odds ratios with their confidence intervals
cbind(OR, CI)
```

Present and interpret a summary of the fitted model.

Present and interpret the coefficients of the model as odds ratios and provide confidence intervals for them.

Interpret the results and compare them to your previously stated hypothesis.



### Explore predictive power

Using the variables which, according to your logistic regression model, had a statistical relationship with high/low alcohol consumption, explore the predictive power of you model. Provide a 2x2 cross tabulation of predictions versus the actual values and optionally display a graphic visualizing both the actual values and the predictions. Compute the total proportion of inaccurately classified individuals (= the training error) and comment on all the results. Compare the performance of the model with performance achieved by some simple guessing strategy. (0-3 points)

Here is how my current model is performing:

```{r, echo=FALSE}
# predictions versus the actual values
    
# predict() the probability of high_use
probabilities <- predict(m, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = probabilities > 0.5 )

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction) %>% 
prop.table() %>% 
addmargins
```

My hypothesis was pretty bad and it is proven by this data. Overall, nearly 95 % of the predictions are wrong! At least my faith in humanity is now restored and some kind of easy stereotypes can be ignored (as they already were).

The variables that actually work from my model are here:

```{r, echo=FALSE, message=FALSE}
# new model
m2 <- glm(high_use ~ absences, data = alc, family = "binomial")

# compute odds ratios (OR)
OR2 <- coef(m2) %>% exp

# compute confidence intervals (CI)
CI2 <- confint(m2) %>% exp


# predict() the probability of high_use
probabilities2 <- predict(m2, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability2 = probabilities2)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction2 = probabilities2 > 0.5 )

# print out a summary of the model
summary(m2)

# print out the coefficients of the model
coef(m2)

# print out the odds ratios with their confidence intervals
cbind(OR2, CI2)

```


I know, it is only one variable...but it seems that students who missed class regularly are more likely to be at risk of being high alcohol consumers. 

Let's see how these two models compare with a likelihood ratio test:

```{r, echo=FALSE}
anova(m, m2, test="LRT")
```


```{r, echo=FALSE}
# initialize a plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = probability2, y = high_use, col= prediction2)) + xlab("probability")

# define the geom as points and draw the plot
g + geom_point() 

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction2) %>% 
prop.table() %>% 
addmargins

```

### Appologies

I am sorry if you have to correct this work, it is done last minute and on the fly. Bad timing to realise that I hadn't understood/needed to study more regarding the topic of this week's matterial. Good luck with your assignment. :)